---
title: "Week 12 - Arranging the data"
subtitle: "CHC9012 | National Taiwan Normal University"
format: html
---

```{=html}
<style>
  
  /* Tabset Styling */
  .nav-tabs {
    border-bottom: 2px solid #ecf0f1;
    margin-top: 1em;
    margin-bottom: 30px;
  }
.nav-tabs .nav-link {
  font-weight: 700;
  font-size: 1.1em;
  color: #95a5a6;
    border: none;
  padding: 15px 25px;
  transition: all 0.3s ease;
}
.nav-tabs .nav-link:hover {
  color: #2c3e50;
    border: none;
}
.nav-tabs .nav-link.active {
  color: #3498db !important;
    background: none;
  border: none;
  border-bottom: 4px solid #3498db;
}

/* Stunning Headers */
  
  h2 {
    margin-top: 1.8em;
    color: #2980b9;
      font-weight: 600;
  }
h3 {
  margin-top: 1.5em;
  color: #2c3e50;
    font-weight: 600;
}

/* Floating Images with Animations */
  img {
    border-radius: 12px;
    box-shadow: 0 8px 25px rgba(0,0,0,0.06);
    transition: transform 0.4s cubic-bezier(0.175, 0.885, 0.32, 1.275), box-shadow 0.4s ease;
    margin: 2rem auto;
    display: block;
  }
img:hover {
  transform: translateY(-6px) scale(1.02);
  box-shadow: 0 15px 35px rgba(52,152,219,0.2);
}

/* Elegant Gradient Divider */
  hr.rounded {
    border: 0;
    height: 2px;
    background-image: linear-gradient(to right, rgba(0,0,0,0), rgba(52,152,219,0.5), rgba(0,0,0,0));
    margin: 4rem 0;
  }

/* Pro Code Styling */
  code {
    font-family: 'JetBrains Mono', monospace;
    font-size: 0.9em;
    color: #e74c3c;
      background-color: #fdf2f0;
      padding: 3px 6px;
    border-radius: 5px;
  }
pre code {
  color: #333;
    background-color: transparent;
}

/* Clean Tables */
  table {
    width: 100%;
    border-collapse: separate;
    border-spacing: 0;
    border-radius: 12px;
    overflow: hidden;
    box-shadow: 0 5px 15px rgba(0,0,0,0.05);
    margin: 2rem 0;
  }
th {
  background-color: #3498db;
    color: white;
  text-transform: uppercase;
  letter-spacing: 1px;
  padding: 15px;
  font-weight: 600;
}
td {
  background-color: white;
  padding: 15px;
  border-bottom: 1px solid #ecf0f1;
}

/* Emphasized bolding */
  b, strong {
    color: #2c3e50;
      font-weight: 700;
  }
  
  .panel-tabset .nav-tabs, 
  .panel-tabset .nav-pills {
    display: flex !important;
    flex-wrap: nowrap !important;
    overflow-x: auto !important;
    overflow-y: hidden !important;
    -webkit-overflow-scrolling: touch; /* Smooth scrolling on mobile */
    border-bottom: 1px solid #dee2e6; /* Keeps the line consistent */
  }

  /* Optional: Hide the scrollbar for a cleaner look while keeping the functionality */
  .panel-tabset .nav-tabs::-webkit-scrollbar,
  .panel-tabset .nav-pills::-webkit-scrollbar {
    display: none;
  }

  .panel-tabset .nav-item {
    white-space: nowrap; /* Prevents text inside tabs from breaking into two lines */
  }
</style>
```

## Overview

This week, we will learn how to arrange the dataset such that it is ready for further exploration.

::::: panel-tabset
## 🎯 Goals

**Prepare your computer:**

-   **Arrangements:** How to prepare your data for further analyses?
-   **Checks:** How to check that all the data are in order?

## 💻 Necessary codes and dataset

**A brief summary of the essential elements:**

-  `load(...)`: Imports a dataset or workspace previously saved in .RData format.
-  `read.csv(...)`: Imports external data, such as your stimulus material list, from a .csv file.
-  `<- NULL`: Used to remove unnecessary columns from the dataframe.
-  `names(...)`: Used to view or rename the column headers of the dataset.
-  `filter(...)`: Keeps only the rows that meet specific conditions (e.g., keeping only "Experiment" trials).
-  `ifelse(...)`: Creates a new value based on a condition.
-  `spread(...)`: Reshapes data from long to wide format.
-  `left_join(...)`: Merges two datasets together based on common identifier columns (like ItemNo).
-  `as.factor(...)` / `as.numeric(...)`: Transforms a variable into a category (factor) or a mathematical number (numeric).
-  `ezDesign(...)`: Generates a heatmap to visually verify that the experimental design is balanced and no data is missing.
-  `save(...)`: Exports your final, prepared dataset into an .RData file for future analysis.

<br>

::: {.text-center .mb-3}
<a href="https://github.com/aymeric-collart/CHC9012/blob/main/course-materials/Files/results_FAKE_FOR_CHC9012.csv" class="btn btn-outline-primary btn-sm"> <i class="bi bi-download"></i> Download the dataset for the script [GIVEN LATER]</a>
:::

## 📝 Homework

**Please complete the following tasks before our next meeting:**

-  [  ] Download the Markdown script
-  [  ] Annotate each line of the script
-  [  ] Knit into a PDF document, and upload on Moodle
:::::

::: {.text-center .mb-4}
<a href="week12_slides.html" class="btn btn-outline-primary" target="_blank"> <i class="bi bi-window-stack"></i> View the slides of this week's tutorial</a>
:::

<br>

# 💭 Part 1: Plan what you need to do

::: {.callout-tip icon="true"}
## ✍️ Preparatory activity

We should never forget that coding is a tool to achieve a task. For our case, the main task is to **prepare the dataset for further analyses**, which means:

-  Checking that it is complete and that nothing is missing,
-  Deleting the columns/rows that are not necessary, and 
-  Transforming (but not *altering*) the data
:::

The best way is to look at the data we just imported into RStudio, and take notes about what needs to be done. I personally take physical notes on flying papers for this step.

```{r}
#| code-fold: true
#| code-summary: "Show the code to import and show the dataset"
#| message: false
#| warning: false

read.pcibex <- function(filepath, auto.colnames=TRUE, fun.col=function(col,cols){cols[cols==col]<-paste(col,"Ibex",sep=".");return(cols)}) {
  n.cols <- max(count.fields(filepath,sep=",",quote=NULL),na.rm=TRUE)
  if (auto.colnames){
    cols <- c()
    con <- file(filepath, "r")
    while ( TRUE ) {
      line <- readLines(con, n = 1, warn=FALSE)
      if ( length(line) == 0) {
        break
      }
      m <- regmatches(line,regexec("^# (\\d+)\\. (.+)\\.$",line))[[1]]
      if (length(m) == 3) {
        index <- as.numeric(m[2])
        value <- m[3]
        if (is.function(fun.col)){
         cols <- fun.col(value,cols)
        }
        cols[index] <- value
        if (index == n.cols){
          break
        }
      }
    }
    close(con)
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=cols))
  }
  else{
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=seq(1:n.cols)))
  }
}

data <- read.pcibex("../course-materials/Files/results_FAKE_FOR_CHC9012.csv")

rmarkdown::paged_table(data, options = list(rows.print = 50))
```

This is what the results table looks like. Let's have a look at its columns, 18 in this file. There are two types: (a) PCIbex's default columns, and (b) the columns that we asked PCIbex to log. 

::::: panel-tabset
## Default columns from PCIbex

-  `Results.reception.time`: Time in the PCIbex server when the results were received. This can be used as the Participant ID, assuming that there are no two participants who submitted the results at the exact same time.
-  `MD5.hash.of.participant.s.IP.address`: Scripted IP address of the machine used by the participant.
-  `Controller.name`: Which controller was used to code the experiment.
-  `Order.number.of.item`: The order of presentation of the whole questionnaire, starting from the very first page.
-  `Inner.element.number`: Whether specific numbers were assigned to the element.
-  `Label`: The name given to the webpage in PCIbex (for example, `Welcome page`, `Instructions`, `Exeriment`, etc.).
-  `Latin.Square.Group`
-  `PennElementType`: The PCIbex element used for the action (`TextInput`, `Scale`, `Dropdown`, etc.).
-  `Parameter`: Breakdown of the elements used on a same page.
-  `Value`: What the participant wrote or selectedm or when an element *starts* or *ends*.
-  `EventTime`: Time point at which an action occured (the time corresponds to the PCIbex server)
-  `Comments`

## Columns that we manually logged

-  `ItemNo`: The ID of the item (here, sentence)
-  `Condition`: The experimental condition
-  `Group`
-  `UserAge`
-  `UserDegree`
:::::

::: {.callout-tip icon="true"}
## ✍️ Preparatory activity, again

Look closely to the results file. What needs to be done?
:::

::::: panel-tabset
## Missing data
-  [ ] Calculate the reaction time (so far, only the event time)
-  [ ] The actual sentences are missing, need to add them based on `ItemNo`
-  [ ] No information about the consent box, but this cannot be retrieved


## Unnecessary data

-  [ ] Columns containing the same or no value (`Controller.name`, `Inner.element.number`, `Latin.Square.Group`, `Comments`, for example)
-  [ ] Keep only the lines where `Label` is "Experiment"

## Data to transform

-  [ ] Columns to rename, to be more concise and explicit (`Results.reception.time`, `MD5.hash.of.participant.s.IP.address`)
-  [ ] Change the nature of the variable (`UserAge` to numerical value, `ItemNo` to a group/factor, etc.)

:::::

<hr>

# 💻 Part 2: Arrange the dataset with RStudio, according to the plan

You can find below how to use RStudio to perform the transformations listed above.

::::: {.panel-tabset .nav-pills .tabset-fade group="pipeline"}
## 2.1 Remove unnecessary columns

-  We need to mention which column from which dataset. In R language, "the column called *Controller.name* from the dataset called *data*" is written using the `$` sign (`data$Controller.name`)
-  We assign `NULL` to this column to delete it.

```{r}
data$Controller.name <- NULL
data$Inner.element.number <- NULL
data$Latin.Square.Group <- NULL
data$Comments <- NULL
```


## 2.2 Rename columns

You can have a glimpse of the names of the columns using the `names(...)` function.

```{r}
names(data)
```

Each name is assigned to a numerical index. For instance, "[1]" corresponds to "Results.reception.time". We can call the name of one specific column as below.

```{r}
names(data)[1]
```

Once we have the index of columns, we just need to assign new names to them.

```{r}
names(data)[1] <- "ID_1_TimeReceived"
names(data)[2] <- "ID_2_IP"

names(data)
```

## 2.3 Select the rows we need

Based on our notes, we need to filter to have only the rows where the `Label` is `Experiment`. 

-  To do so, we use the `filter()` function from the `dplyr` package.
-  We use the `%>%` to say that "This is the `Label` column from the dataset called `data`".
-  The `==` sign means that we *want* the `Experiment` rows.
-  We store that into a new dataset, just in case. 

```{r}
#| message: false
#| warning: false

library(dplyr)

data2 <- data %>%
  filter(Label == "Experiment")
```

```{r}
#| echo: false

rmarkdown::paged_table(data2, options = list(rows.print = 10))
```

We can see that there are unnecessary `_Header_` rows under the `Parameter` column, and we would like to remove them. We can use the same logic, but notice the use of `!=`, meaning "we *don't want* these rows".

```{r}
data2 <- data2 %>%
  filter(Parameter != "_Header_")
```

```{r}
#| echo: false

rmarkdown::paged_table(data2, options = list(rows.print = 10))
```

## 2.4 Calculate the reaction time

::: {.callout-tip icon="false"}
## 📖 Defining reaction time

In this experiment, `Reaction time` = `Time when the judgment is made` - `Time when the sentence appeared`

Calculating the reaction time from PCIbex data requires several steps.
:::

Let's have a look at our data:

```{r}
#| echo: false
data_show <- data2 %>%
  select(Parameter, Value, EventTime, ItemNo)
rmarkdown::paged_table(data_show, options = list(rows.print = 6))
```

Here, `Reaction time` = "EventTime *when* `Parameter` = `Choice`" - "EventTime *when* `Value` = `Start`". So we need to select these rows first. We stored in a temporary dataset that we call `RT_Prep1`.

```{r}
RT_Prep1 <- data2 %>%
  filter(Parameter == "Choice" | Value == "Start")
```

Under the `Value` column, we have the time for `Start`, but not for `Choice`, which is found in another column. We want to have everything under the same column. So we tell R that when in the `Value` column, the word is not `Start`, it writes `Choice` instead. Then, we delete columns when the two rows are different -- we want to keep the differences minimal.

```{r}
RT_Prep1$Value <- ifelse(RT_Prep1$Value == "Start",
                         "Start",
                         "Choice")

RT_Prep1$PennElementType <- NULL
RT_Prep1$PennElementName <- NULL
RT_Prep1$Parameter <- NULL
```

```{r}
#| echo: false

RT_Show <- RT_Prep1 %>%
  select(Value, EventTime, ItemNo)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

To do the subtraction, we want to have the `EventTime` for `Start` and `Choice` in different columns. We use the `spread()` function from the `tidyr` library.

```{r}
library(tidyr)
RT_Prep2 <- spread(data = RT_Prep1, Value, EventTime)

```

```{r}
#| echo: false

RT_Show <- RT_Prep2 %>%
  select(ItemNo, Start, Choice)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

Now we just have to tell R that we are dealing with numbers, and do the subtraction.

```{r}
RT_Prep2$Choice <- as.numeric(RT_Prep2$Choice)
RT_Prep2$Start <- as.numeric(RT_Prep2$Start)
RT_Prep2$RT <- RT_Prep2$Choice - RT_Prep2$Start
```

```{r}
#| echo: false

RT_Show <- RT_Prep2 %>%
  select(ItemNo, Start, Choice, RT)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

::: {.callout-tip icon="false"}
## 💭 Last step

We have two datasets:

-  `data2`, with the judgment data, but ***not*** the reaction time,
-  `RT_Prep2`, the temporary dataset with the reaction time, but ***not*** the judgment data.

We just need to join them based on their common minimal value: 1 different trial per participant.
:::

First, we select the row where the judgment is given.

```{r}
data3 <- data2 %>%
  filter(Parameter %in% "Choice")
```

Finally, we use the `left_join()` function to join these two datasets based on (a) the participant ID, and (b) the Item index.

```{r}
data3 <- data3 %>%
  left_join(RT_Prep2 %>% select(ID_1_TimeReceived, ID_2_IP, ItemNo, RT), 
            by = c("ID_1_TimeReceived", "ID_2_IP", "ItemNo"))
```
```{r}
#| echo: false

RT_Show <- data3 %>%
  select(ID_1_TimeReceived, ItemNo, Value, RT, Condition)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 25))
```


## 2.5 Change the nature of the variables

Let's look again at our data:
```{r}
str(data3)
```

Some variables correspond to ***groups*** (or categories), so we used the `as.factor()` function to group them.

```{r}
data3$ID_1_TimeReceived <- as.factor(data3$ID_1_TimeReceived)
data3$ID_2_IP <- as.factor(data3$ID_2_IP)
data3$ItemNo <- as.factor(data3$ItemNo)
data3$Condition <- as.factor(data3$Condition)
data3$Group <- as.factor(data3$Group)
data3$UserDegree <- as.factor(data3$UserDegree)
```

Some are just numerical value, and so we use the `as.numeric()` function.

```{r}
data3$Order.number.of.item <- as.numeric(data3$Order.number.of.item)
data3$Value <- as.numeric(data3$Value)
data3$UserAge <- as.numeric(data3$UserAge)
data3$RT <- as.numeric(data3$RT)
```

Now this is our data:
```{r}
str(data3)
```


## 2.6 Another round of removing unnecessary columns

We made use of some columns when calculating the reaction time, but we do not need them anymore, and they are redudant. So here is another round of column removal!

```{r}
str(data3)

data3$Label <- NULL
data3$PennElementType <- NULL
data3$PennElementName <- NULL
data3$Parameter <- NULL
data3$EventTime <- NULL

str(data3)
```

## 2.7 Add information about the items

Finally, we realized that even if we have enough information for each trial to conduct further analyses, we do not know what sentence they correspond to. This is surely because we forgot to `.log()` them in the PCIbex script. But we can repair this mistake:

-  We just need to import the table with the material information (used on the PCIbex platform)
-  We use the `left_join()` function as previously to join these two datasets based on the index of the items (here, `ItemNo`)


```{r}
material <- read.csv("~/Documents/CHC9012_Spring2026/PreprocessingAnalysesR/material.csv")
material$ItemNo <- as.factor(material$ItemNo)

data3 <- data3 %>%
  left_join(material %>% select(ItemNo, Sentence), 
            by = "ItemNo")

data3$Sentence <- as.factor(data3$Sentence)
str(data3)
```

:::::

<hr>

# 💾 Part 3: Check and save the data

::: {.callout-tip icon="true"}
## 💭 Checking process

We want to make sure that we did not corrupt our data when arranging them for further analyses. Otherwise, we can run further analyses and have results, but they will have no value!

We will mostly use the `ezDesign()` function from the `ez` package.
:::

::::: panel-tabset

## 3.1 Check the data

```{r}
#| message: false
#| warning: false
library(ez)
```

```{r}
# Same number of item per condition for each participant? (should be '15')
ezDesign(data3, ID_1_TimeReceived, Condition)

# Which participants share the same ID? (hint that this could be the same participant filling the questionnaire twice)
ezDesign(data3, ID_1_TimeReceived, ID_2_IP)

# Number of times an item appeared by participant (should be '1')
ezDesign(data3, ID_1_TimeReceived, ItemNo)

# Were the items well dispatched in the two lists?
ezDesign(data3, ItemNo, Group)

# Were the participants well dispatched in the two lists? (should be '30')
ezDesign(data3, ID_1_TimeReceived, Group)
```

## 3.2 Save the prepared dataset

```{r}
save(data3,
     file = "CHC9012_PCIbexDataPrepared.RData")
```

::::: 

<hr>

::: {.callout-important icon="true"}
## 📝 Assignment for next week

Please complete the following tasks before our next meeting:

-  [  ] Download the Markdown script
-  [  ] Annotate each line of the script
-  [  ] Knit into a PDF document, and upload on Moodle
:::