---
title: "Week 14 - Visualizing the data"
subtitle: "CHC9012 | National Taiwan Normal University"
format: 
  revealjs:
    theme: serif
    transition: slide
    scrollable: true
    slide-number: true
    chalkboard: true
    width: 1280
    height: 720
---

## 🎯 Goals

**Visualize your results:**

::: {.fragment .fade-up}
-   What types of figures to select?
:::

::: {.fragment .fade-up}
-   How to plot multiple information in one figure?
:::


---

## 💻 Dataset and script

Please download the files for the content of Week 13, and place them in your `PreprocessingAnalysesR` folder.

::: {style="text-align: center; margin-top: 2em;"}

<a href="https://github.com/aymeric-collart/CHC9012/blob/main/course-materials/Files/_Internal/CHC9012_PCIbexDataPrepared.RData?raw=true" 
   style="display: inline-block; padding: 15px 30px; background-color: #447099; color: white; text-decoration: none; border-radius: 10px; font-weight: bold; margin-bottom: 25px; box-shadow: 0 4px 6px rgba(0,0,0,0.1);">
   📥 Download the dataset for the script
</a>

<a href="https://aymeric-collart.github.io/CHC9012/course-materials/Files/_Internal/DataVisualization.Rmd" 
   download="DataVisualization.Rmd" 
   style="display: inline-block; padding: 15px 30px; border: 2px solid #447099; color: #447099; text-decoration: none; border-radius: 10px; font-weight: bold;">
   📥 Download the R Markdown script
</a>

:::
::: {.fragment .fade-up}
<br>
💻 Please start `RStudio` and select the project called `PreprocessingAnalysesR`.
:::


# 👀 Part 1: Why using figures, and what kinds of figures?

## Why plotting the data?

**Why using a figure when we already have the table of results?**

<div class="pulse-box">
  ⏳ **Activity:** Please spend 10 minutes to brainstorm this question.
</div>

<style>
.pulse-box {
  background-color: #eef2f7;
  border-left: 5px solid #0d6efd;
  padding: 20px;
  margin-top: 40px;
  border-radius: 4px;
  font-size: 2.5rem; 
  font-weight: bold;
  text-align: center;
  animation: pulse-animation 2s infinite;
}

@keyframes pulse-animation {
  0% { transform: scale(1); box-shadow: 0 0 0 0 rgba(13, 110, 253, 0.4); }
  70% { transform: scale(1.02); box-shadow: 0 0 0 10px rgba(13, 110, 253, 0); }
  100% { transform: scale(1); box-shadow: 0 0 0 0 rgba(13, 110, 253, 0); }
}
</style>

---

## Why plotting the data?

**Why using a figure when we already have the table of results?**

-   **Nature of human cognition.** Figures = more attractive than tables and texts.

-   **Eyes sharper than we think.** It is sometimes easier to observe tendencies with a figure.

# Introduction to most common types of figures in experimental linguistics

## Box Plots for *Data Distribution*

::: {.columns}

::: {.column width="45%"}

::: {.fragment .fade-up .callout-note appearance="simple"}
###  What it does:
Visualizes the spread, median, quartiles, and outliers of a continuous numerical variable across different categorical groups.
:::

<br>

::: {.fragment .fade-up .callout-note appearance="simple"}
###  Core question
**How does data distribution compare between groups?** *(e.g., Which Iris species has the longest petals?)*
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.5
#| fig-height: 5.5

library(ggplot2)

ggplot(iris, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    title = "Petal length distribution by species", 
    y = "Petal length (in cm)",
    x = "Species"
  ) +
  theme_minimal(base_size = 16) + # Bumps text size up so it is readable on the projector
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 18)
  )
```
:::

:::

---

## Bar plots for *categorical counts*

::: {.columns}

::: {.column width="45%"}

::: {.fragment .fade-up .callout-note appearance="simple"}
###  What it does:
Displays counts or frequencies of distinct categories.
:::

<br>

::: {.fragment .fade-up .callout-note appearance="simple"}
###  Core question
Which categories are the most and least common in our dataset?
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.5
#| fig-height: 5.5

ggplot(mpg, aes(x = class)) +
  geom_bar(fill = "steelblue", alpha = 0.85, width = 0.7) +
  labs(title = "Count of vehicles by class", x = "Vehicle type", y = "Count") +
  theme_minimal()
```
:::

:::

---

## Line charts for *continuous measurement, generally over time*

::: {.columns}

::: {.column width="45%"}

::: {.fragment .fade-up .callout-note appearance="simple"}
###  What it does:
Connects individual data points chronologically or sequentially to reveal trajectories, spikes, or downward dips.
:::

<br>

::: {.fragment .fade-up .callout-note appearance="simple"}
###  Core question
How is our metric moving over a continuous sequence?
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.5
#| fig-height: 5.5

ggplot(economics, aes(x = date, y = psavert)) +
  geom_line(color = "firebrick", linewidth = 1) +
  labs(title = "Personal savings rate trend", x = "Year", y = "Savings rate (in %)") +
  theme_minimal()
```
:::

:::

---

## Scatter plots for *relationships & correlations*

::: {.columns}

::: {.column width="45%"}

::: {.fragment .fade-up .callout-note appearance="simple"}
###  What it does:
Plots individual data points at the intersection of two continuous numeric variables to display positive, negative, or non-linear correlations.
:::

<br>

::: {.fragment .fade-up .callout-note appearance="simple"}
###  Core question
Does variable X increase, decrease, or remain unchanged as variable Y changes?
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.5
#| fig-height: 5.5

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 2.5, alpha = 0.8) +
  labs(title = "Sepal length vs. Sepal width", x = "Length (in cm)", y = "Width (in cm)") +
  theme_minimal()
```
:::

:::

---

## Combined plots for *more complex data*

::: {.columns}

::: {.column width="45%"}

::: {.fragment .fade-up .callout-note appearance="simple"}
###  What it does:
Plots two types of data at the same time.
<br><br>
Here, a **boxplot** with an additional **scatter plot**.
:::
:::

::: {.column width="55%"}
```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6.5
#| fig-height: 5.5

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  # 1. Plot the underlying summary structural distribution
  geom_boxplot(alpha = 0.4, outlier.shape = NA, width = 0.5) + 
  
  # 2. Layer the raw data points on top (jittered slightly horizontally so they don't stack up)
  geom_jitter(aes(color = Species), width = 0.2, size = 2, alpha = 0.7) +
  
  labs(
    title = "Sepal length by species",
    subtitle = "Combining box plots with jittered data points to avoid hiding sample distribution",
    y = "Sepal length (in cm)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
:::

:::

# Part 2: A step-by-step guide to plot figures with R


# 📝 Assignment for next week

::: {.callout-important icon="true"}
Please annotate the Markdown file from this week, knit to a PDF file, and upload on Moodle!
:::
