---
title: "Week 13 - Exploring the data"
subtitle: "CHC9012 | National Taiwan Normal University"
format: 
  revealjs:
    theme: serif
    transition: slide
    scrollable: true
    slide-number: true
    chalkboard: true
    width: 1280
    height: 720
---

## 🎯 Goals

**Explore your data:**

::: {.fragment .fade-up}
-   **Non-valid data:** How to identify and remove non-valid data?
:::

::: {.fragment .fade-up}
-   **Analyses:** How to analyze and further explore your data?
:::


---

## 💻 Dataset and script

Please download the files for the content of Week 13, and place them in your `PreprocessingAnalysesR` folder.

::: {style="text-align: center; margin-top: 2em;"}

<a href="https://github.com/aymeric-collart/CHC9012/blob/main/course-materials/Files/_Internal/CHC9012_PCIbexDataPrepared.RData?raw=true" 
   style="display: inline-block; padding: 15px 30px; background-color: #447099; color: white; text-decoration: none; border-radius: 10px; font-weight: bold; margin-bottom: 25px; box-shadow: 0 4px 6px rgba(0,0,0,0.1);">
   📥 Download the dataset for the script
</a>

<a href="https://aymeric-collart.github.io/CHC9012/course-materials/Files/_Internal/DataExploration.Rmd" 
   download="DataExploration.Rmd" 
   style="display: inline-block; padding: 15px 30px; border: 2px solid #447099; color: #447099; text-decoration: none; border-radius: 10px; font-weight: bold;">
   📥 Download the R Markdown script
</a>

:::
::: {.fragment .fade-up}
<br>
💻 Please start `RStudio` and select the project called `PreprocessingAnalysesR`.
:::

---

## 💻 Preparations for the analyses

```r
load(file = "CHC9012_PCIbexDataPrepared.RData")

library(ez)
library(dplyr)
library(Rmisc)
```

```{r}
#| message: false
#| warning: false

load(file = "CHC9012_PCIbexDataPrepared.RData")

library(ez)
library(dplyr)
library(Rmisc)
```


# 👀 Part 1: Have an overview of the results

---

## Overall results (1/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$Value ~ data3$Condition)
```

```{r}
plot(data3$Value ~ data3$Condition)
```
:::

::: {.column width="30%"}

**Distribution of the ratings across conditions**

<br><br>
Please briefly describe the patterns of results.
:::

::::

---

## Overall results (2/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$RT ~ data3$Condition)
```

```{r}
plot(data3$RT ~ data3$Condition)
```
:::

::: {.column width="30%"}

**Distribution of the ratings across conditions**

<br><br>
Please briefly describe the patterns of results.
:::

::::

---

## Overall results (2/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$RT ~ data3$Condition, outline = FALSE)
```

```{r}
plot(data3$RT ~ data3$Condition, outline = FALSE)
```
:::

::: {.column width="30%"}

**Distribution of the ratings across conditions**

<br><br>
Please briefly describe the patterns of results.
:::

::::

---

## Overall results by participant (1/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$Value ~ data3$ID_1_TimeReceived)
```

```{r}
plot(data3$Value ~ data3$ID_1_TimeReceived)
```
:::

::: {.column width="30%"}

**Distribution of the ratings by participants**

<br><br>
Are there many very deviant participants?
:::

::::

---

## Overall results by participant (2/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$RT ~ data3$ID_1_TimeReceived, outline = FALSE)
```

```{r}
plot(data3$RT ~ data3$ID_1_TimeReceived, outline = FALSE)
```
:::

::: {.column width="30%"}

**Distribution of the ratings by participants**

<br><br>
Are there many very deviant participants?
:::

::::

---

## Overall results by item (1/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$Value ~ data3$ItemNo)
```

```{r}
plot(data3$Value ~ data3$ItemNo)
```
:::

::: {.column width="30%"}

**Distribution of the ratings by participants**

<br><br>
Are there many very deviant items?
:::

::::

---


## Overall results by item (2/2)


:::: {.columns}

::: {.column width="70%"}

```r
plot(data3$RT ~ data3$ItemNo, outline = FALSE)
```

```{r}
plot(data3$RT ~ data3$ItemNo, outline = FALSE)
```
:::

::: {.column width="30%"}

**Distribution of the ratings by participants**

<br><br>
Are there many very deviant items?
:::

::::

---

## How to look at the background of the participants?

::: {.fragment .fade-up}

**Number of participants**

```r
length(levels(data3$ID_1_TimeReceived))
```
```{r}
length(levels(data3$ID_1_TimeReceived))
```

:::

::: {.fragment .fade-up}

**Age**

```r
summary(data3$UserAge)
```

```{r}
summary(data3$UserAge)
```


:::

::: {.fragment .fade-up}

**Level of education**

```r
DegreePart <- subset(data3, select = c(ID_1_TimeReceived, UserDegree))
DegreePart <- unique(DegreePart)
table(DegreePart$UserDegree)
```

```{r}
DegreePart <- subset(data3, select = c(ID_1_TimeReceived, UserDegree))
DegreePart <- unique(DegreePart)
table(DegreePart$UserDegree)
```

:::

---

## Summary

::: {.fragment .fade-up}
-   **Overall pattern:** Grammatical sentences more accepted that ungrammatical ones, but not very obvious
:::

::: {.fragment .fade-up}
-   **Non-valid data:** Many participants with a strange behavior
:::

# 🤯 Part 2: Identify and remove outliers

---

## 2.1. Identify and remove non-valid participants


::: {.callout-important icon="false"}
## ⚠️ Be careful

We need to be careful when removing data! We need **converging evidence** and ***justify our choice***.
:::

::: {.fragment .fade-up}
Different types of evidence to identify non-valid participants:
:::
::: {.fragment .fade-up}
-  Based on abnormal mean reaction time
:::
::: {.fragment .fade-up}
-  Based on shared IPs (shared device)
:::
::: {.fragment .fade-up}
-  Based on the observation of specific participants
:::

---

## Abnormal RT: Overall observations

:::: {.columns}

::: {.column width="65%"}
```r
plot(data3$RT ~ data3$ID_1_TimeReceived, outline = FALSE)
```
```{r}
plot(data3$RT ~ data3$ID_1_TimeReceived, outline = FALSE)
```
:::

::: {.column width="35%"}

```r
PartMeanRT <- aggregate(RT ~ ID_1_TimeReceived,
                        data = data3,
                        FUN = mean)

print(PartMeanRT)
```

```{r}
PartMeanRT <- aggregate(RT ~ ID_1_TimeReceived,
                        data = data3,
                        FUN = mean)

print(PartMeanRT)
```
:::

::::

---

## Abnormal RT: Focus on the weird behavior

:::: {.columns}

::: {.column width="65%"}

```r
data_QuickPart <- subset(PartMeanRT, RT < 1000)
data_QuickPart_FullData <- subset(data3,
                                  ID_1_TimeReceived %in% data_QuickPart$ID_1_TimeReceived)

plot(data_QuickPart_FullData$RT ~ data_QuickPart_FullData$ID_1_TimeReceived, outline = FALSE)
```

```{r}
data_QuickPart <- subset(PartMeanRT, RT < 1000)
data_QuickPart_FullData <- subset(data3,
                                  ID_1_TimeReceived %in% data_QuickPart$ID_1_TimeReceived)

plot(data_QuickPart_FullData$RT ~ data_QuickPart_FullData$ID_1_TimeReceived, outline = FALSE)
```

:::

::: {.column width="35%"}

```r
plot(data_QuickPart_FullData$Value ~ data_QuickPart_FullData$Condition)
```

```{r}
plot(data_QuickPart_FullData$Value ~ data_QuickPart_FullData$Condition)
```

:::

::::

---

## Abnormal RT: Focus on the human-like behavior

:::: {.columns}

::: {.column width="65%"}

```r
data_OtherPart <- subset(PartMeanRT, RT > 1000)
data_OtherPart_FullData <- droplevels(subset(data3,
                                             ID_1_TimeReceived %in% data_OtherPart$ID_1_TimeReceived))

plot(data_OtherPart_FullData$RT ~ data_OtherPart_FullData$ID_1_TimeReceived, outline = FALSE)
```

```{r}
data_OtherPart <- subset(PartMeanRT, RT > 1000)
data_OtherPart_FullData <- droplevels(subset(data3,
                                             ID_1_TimeReceived %in% data_OtherPart$ID_1_TimeReceived))

plot(data_OtherPart_FullData$RT ~ data_OtherPart_FullData$ID_1_TimeReceived, outline = FALSE)
```

:::
::: {.column width="35%"}

```r
plot(data_OtherPart_FullData$Value ~ data_OtherPart_FullData$Condition)
```

```{r}
plot(data_OtherPart_FullData$Value ~ data_OtherPart_FullData$Condition)
```


:::
::::


## Shared IPs (shared device)

:::: {.columns}

::: {.column width="65%"}

```{r}
ezDesign(data_QuickPart_FullData, ID_1_TimeReceived, ID_2_IP)
```

:::
::: {.column width="35%"}

**Weird RTs**

```r
ezDesign(data_QuickPart_FullData,
ID_1_TimeReceived,
ID_2_IP)
```
:::
::::

:::: {.columns}
::: {.column width="65%"}

```{r}
ezDesign(data_OtherPart_FullData, ID_1_TimeReceived, ID_2_IP)
```

:::
::: {.column width="35%"}

**Human-like behavior**

```r
ezDesign(data_OtherPart_FullData,
ID_1_TimeReceived,
ID_2_IP)
```
:::
::::

## Look at specific participants

**Code to extract the data of one participant (here, Participant #9)**
```r
IndPart <- subset(data_OtherPart_FullData,
                  ID_1_TimeReceived == levels(data_OtherPart_FullData$ID_1_TimeReceived)[9])
```
```{r}
IndPart <- subset(data_OtherPart_FullData,
                  ID_1_TimeReceived == levels(data_OtherPart_FullData$ID_1_TimeReceived)[9])
```

<br>

:::: {.columns}

::: {.column width="50%"}
```r
plot(IndPart$Value ~ IndPart$Condition)
```
```{r}
plot(IndPart$Value ~ IndPart$Condition)
```

:::

::: {.column width="50%"}
```r
plot(IndPart$RT ~ IndPart$Condition)
```
```{r}
plot(IndPart$RT ~ IndPart$Condition)
```
:::

::::

---

## Summary

::: {.fragment .fade-up}
-   **Converging evidence** to say that one group's behavior is very different, indicating that they **did not pay attention to the task**
:::

::: {.fragment .fade-up}
-   **Further analyses:** Inclusion of participants with the human-like behavior
:::


## 2.2. Identify and remove non-valid trials

```{tikz}
#| echo: false
\usetikzlibrary{arrows.meta, positioning, fit, backgrounds}
\begin{tikzpicture}[
    base/.style={rectangle, draw, thick, minimum width=2.8cm, minimum height=1.2cm, align=center, font=\sffamily\small},
    main/.style={base, fill=blue!10},
    sub/.style={base, fill=white, font=\sffamily\scriptsize, minimum width=2.2cm, minimum height=0.8cm},
    group/.style={draw, thick, dashed, inner sep=8pt, rounded corners, fill=gray!2},
    arrow/.style={-{Stealth[scale=1.2]}, thick}
]

    % Main Path
    \node (b1) [main] {Remove based on \\absolute values};
    \node (b2) [main, right=1.2cm of b1] {Remove based on \\relative values\\(standard deviation)};
    \node (b4) [main, right=1.2cm of b2] {Calculate porportion\\of removed data\\(below or around 5 percent)};

    % Arrows
    \draw [arrow] (b1) -- (b2);
    \draw [arrow] (b2) -- (b4);

\end{tikzpicture}
```


## Remove based on absolute values

```r
data_OtherPart_CleanAbsolute <- subset(data_OtherPart_FullData, RT > 200 & RT < 20000)
```

```{r}
data_OtherPart_CleanAbsolute <- subset(data_OtherPart_FullData, RT > 200 & RT < 20000)
```


## Remove based on relative values (1/2)

```r
plot(data_OtherPart_CleanAbsolute$RT ~ data_OtherPart_CleanAbsolute$ID_1_TimeReceived)
```
```{r}
plot(data_OtherPart_CleanAbsolute$RT ~ data_OtherPart_CleanAbsolute$ID_1_TimeReceived)
```

## Remove based on relative values (2/2)

```r
# Calculate the mean per Participant and Condition
data_OtherPart_CleanAbsolute$m_rt <- ave(data_OtherPart_CleanAbsolute$RT, 
                                         data_OtherPart_CleanAbsolute$ID_1_TimeReceived, 
                                         data_OtherPart_CleanAbsolute$Condition, 
                                         FUN = mean)

# 2. Calculate the SD per Participant and Condition
data_OtherPart_CleanAbsolute$sd_rt <- ave(data_OtherPart_CleanAbsolute$RT, 
                                          data_OtherPart_CleanAbsolute$ID_1_TimeReceived, 
                                          data_OtherPart_CleanAbsolute$Condition, 
                                          FUN = sd)

# 3. Subset the data to remove outliers
data_OtherPart_CleanAbsoluteRelative <- subset(data_OtherPart_CleanAbsolute, 
                                               RT >= (m_rt - 2.5 * sd_rt) & 
                                                 RT <= (m_rt + 2.5 * sd_rt))
```

```{r}
# Calculate the mean per Participant and Condition
data_OtherPart_CleanAbsolute$m_rt <- ave(data_OtherPart_CleanAbsolute$RT, 
                                         data_OtherPart_CleanAbsolute$ID_1_TimeReceived, 
                                         data_OtherPart_CleanAbsolute$Condition, 
                                         FUN = mean)

# 2. Calculate the SD per Participant and Condition
data_OtherPart_CleanAbsolute$sd_rt <- ave(data_OtherPart_CleanAbsolute$RT, 
                                          data_OtherPart_CleanAbsolute$ID_1_TimeReceived, 
                                          data_OtherPart_CleanAbsolute$Condition, 
                                          FUN = sd)

# 3. Subset the data to remove outliers
data_OtherPart_CleanAbsoluteRelative <- subset(data_OtherPart_CleanAbsolute, 
                                               RT >= (m_rt - 2.5 * sd_rt) & 
                                                 RT <= (m_rt + 2.5 * sd_rt))
```


## Calculate the proportion of removed trials


```r
FullData_NbTrials <- nrow(data_OtherPart_FullData)
TrimmedData_NbTrials <- nrow(data_OtherPart_CleanAbsoluteRelative)

# Number of trimmed trials
NbTrimmedTrials <- FullData_NbTrials - TrimmedData_NbTrials
NbTrimmedTrials

# Proportion (%)
NbTrimmedTrials*100/FullData_NbTrials
```

```{r}
FullData_NbTrials <- nrow(data_OtherPart_FullData)
TrimmedData_NbTrials <- nrow(data_OtherPart_CleanAbsoluteRelative)

# Number of trimmed trials
NbTrimmedTrials <- FullData_NbTrials - TrimmedData_NbTrials
NbTrimmedTrials

# Proportion (%)
NbTrimmedTrials*100/FullData_NbTrials
```

The proportion is below 5%, which is acceptable in experimental linguistics.

# 📊 Part 3: Analyses and explorations of the clean results

---

## 3.1: Design-based analyses: Ratings

```r
plot(data_OtherPart_CleanAbsoluteRelative$Value ~ data_OtherPart_CleanAbsoluteRelative$Condition)

summarySE(data_OtherPart_CleanAbsoluteRelative,
          measurevar = "Value",
          groupvars = "Condition")
```

```{r}
plot(data_OtherPart_CleanAbsoluteRelative$Value ~ data_OtherPart_CleanAbsoluteRelative$Condition)

summarySE(data_OtherPart_CleanAbsoluteRelative,
          measurevar = "Value",
          groupvars = "Condition")
```

## 3.1: Design-based analyses: Reaction time

```r
plot(data_OtherPart_CleanAbsoluteRelative$RT ~ data_OtherPart_CleanAbsoluteRelative$Condition)

summarySE(data_OtherPart_CleanAbsoluteRelative,
          measurevar = "RT",
          groupvars = "Condition")
```

```{r}
plot(data_OtherPart_CleanAbsoluteRelative$RT ~ data_OtherPart_CleanAbsoluteRelative$Condition)

summarySE(data_OtherPart_CleanAbsoluteRelative,
          measurevar = "RT",
          groupvars = "Condition")
```

## 3.2: Exploratory analyses

::: {.callout-note icon="true"}
## Why exloring the data?

We observed more variation in the ungrammatical condition. The next step is to understand the source of such variation.

We already did our best with the participants, so it is likely to be **related to the sentences**.
:::

## 3.2: Exploratory analyses, Grammatical condition

```r
data_Grammatical <- droplevels(subset(data_OtherPart_CleanAbsoluteRelative,
                                        Condition == "Grammatical"))

plot(data_Grammatical$Value ~ data_Grammatical$ItemNo)
```
```{r}
data_Grammatical <- droplevels(subset(data_OtherPart_CleanAbsoluteRelative,
                                        Condition == "Grammatical"))

plot(data_Grammatical$Value ~ data_Grammatical$ItemNo)
```

## 3.2: Exploratory analyses, Ungrammatical condition (1/3)

```r
data_Ungrammatical <- droplevels(subset(data_OtherPart_CleanAbsoluteRelative,
                                        Condition == "Ungrammatical"))

plot(data_Ungrammatical$Value ~ data_Ungrammatical$ItemNo)
```

```{r}
data_Ungrammatical <- droplevels(subset(data_OtherPart_CleanAbsoluteRelative,
                                        Condition == "Ungrammatical"))

plot(data_Ungrammatical$Value ~ data_Ungrammatical$ItemNo)
```

## 3.2: Exploratory analyses, Ungrammatical condition (2/3)

```r
# Exploration of divergent ungrammatical sentences
data_Ungrammatical_OutlierValue <- subset(data_Ungrammatical, Value > 4)

data_Ungrammatical_OutlierValue_Summary <- droplevels(subset(data_Ungrammatical_OutlierValue,
                                                  select = c(ID_1_TimeReceived,
                                                             ItemNo,
                                                             Sentence)))
data_Ungrammatical_OutlierValue_Summary <- unique(data_Ungrammatical_OutlierValue_Summary)
data_Ungrammatical_OutlierValue_Overview <- table(data_Ungrammatical_OutlierValue_Summary$ItemNo)

data_Ungrammatical_OutlierValue_Overview
```

```{r}
# Exploration of divergent ungrammatical sentences
data_Ungrammatical_OutlierValue <- subset(data_Ungrammatical, Value > 4)

data_Ungrammatical_OutlierValue_Summary <- droplevels(subset(data_Ungrammatical_OutlierValue,
                                                  select = c(ID_1_TimeReceived,
                                                             ItemNo,
                                                             Sentence)))
data_Ungrammatical_OutlierValue_Summary <- unique(data_Ungrammatical_OutlierValue_Summary)
data_Ungrammatical_OutlierValue_Overview <- table(data_Ungrammatical_OutlierValue_Summary$ItemNo)

data_Ungrammatical_OutlierValue_Overview
```

## 3.2: Exploratory analyses, Ungrammatical condition (3/3)

```r
# Identify which ungrammatical sentences
data_Ungrammatical_OutlierValue_Overview <- as.data.frame(data_Ungrammatical_OutlierValue_Overview)
names(data_Ungrammatical_OutlierValue_Overview)[1] <- "ItemNo"

data_Ungrammatical_OutlierValue_Overview <- droplevels(subset(data_Ungrammatical_OutlierValue_Overview, Freq > 2))

data_Ungrammatical_OutlierValue_Overview <- left_join(data_Ungrammatical_OutlierValue_Overview, 
                                                      data_Ungrammatical, 
                                                      by = "ItemNo")
OutlierUngrammaticalSentences <- unique(data_Ungrammatical_OutlierValue_Overview$Sentence)
OutlierUngrammaticalSentences
```

```{r}
# Identify which ungrammatical sentences
data_Ungrammatical_OutlierValue_Overview <- as.data.frame(data_Ungrammatical_OutlierValue_Overview)
names(data_Ungrammatical_OutlierValue_Overview)[1] <- "ItemNo"

data_Ungrammatical_OutlierValue_Overview <- droplevels(subset(data_Ungrammatical_OutlierValue_Overview, Freq > 2))

data_Ungrammatical_OutlierValue_Overview <- left_join(data_Ungrammatical_OutlierValue_Overview, 
                                                      data_Ungrammatical, 
                                                      by = "ItemNo")
OutlierUngrammaticalSentences <- unique(data_Ungrammatical_OutlierValue_Overview$Sentence)
OutlierUngrammaticalSentences
```

# 📝 Assignment for next week

::: {.callout-important icon="true"}
Please annotate the Markdown file from this week, knit to a PDF file, and upload on Moodle!
:::
