---
title: "Week 12 - Arranging the data"
subtitle: "CHC9012 | National Taiwan Normal University"
format: 
  revealjs:
    theme: serif
    transition: slide
    scrollable: true
    slide-number: true
    chalkboard: true
    width: 1280
    height: 720
---

## 🎯 Goals

**Prepare your dataset:**

::: {.fragment .fade-up}
-   **Arrangements:** How to prepare your data for further analyses?
:::

::: {.fragment .fade-up}
-   **Checks:** How to check that all the data are in order?
:::


---

## 💻 Dataset and script

Please download the files for the content of Week 12, and place them in your `PreprocessingAnalysesR` folder.

::: {style="text-align: center; margin-top: 2em;"}

<a href="https://github.com/aymeric-collart/CHC9012/blob/main/course-materials/Files/_Internal/CHC9012_DataImportedFromPCIbex.RData?raw=true" 
   style="display: inline-block; padding: 15px 30px; background-color: #447099; color: white; text-decoration: none; border-radius: 10px; font-weight: bold; margin-bottom: 25px; box-shadow: 0 4px 6px rgba(0,0,0,0.1);">
   📥 Download the dataset for the script
</a>

<a href="https://aymeric-collart.github.io/CHC9012/course-materials/Files/_Internal/DataPreparationAndCleaning.Rmd" 
   download="DataPreparationAndCleaning.Rmd" 
   style="display: inline-block; padding: 15px 30px; border: 2px solid #447099; color: #447099; text-decoration: none; border-radius: 10px; font-weight: bold;">
   📥 Download the R Markdown script
</a>

:::


# 💭 Part 1: Plan what you need to do

---

## ✍️ Preparatory activity {vertical-align="center"}

::: {.callout-tip icon="true"}
## Instructions


We should never forget that coding is a tool to achieve a task. For our case, the main task is to **prepare the dataset for further analyses**, which means:

-  Checking that it is complete and that nothing is missing,
-  Deleting the columns/rows that are not necessary, and 
-  Transforming (but not *altering*) the data
:::

::: {.fragment .fade-up}
**How to do so?** Two steps:

-  Look at the dataset you just imported
-  Take notes
:::

---

## Step 1: Look at the imported dataset

```{r}
#| code-fold: true
#| code-summary: "Show the code to import and show the dataset (same as last week)"
#| message: false
#| warning: false

read.pcibex <- function(filepath, auto.colnames=TRUE, fun.col=function(col,cols){cols[cols==col]<-paste(col,"Ibex",sep=".");return(cols)}) {
  n.cols <- max(count.fields(filepath,sep=",",quote=NULL),na.rm=TRUE)
  if (auto.colnames){
    cols <- c()
    con <- file(filepath, "r")
    while ( TRUE ) {
      line <- readLines(con, n = 1, warn=FALSE)
      if ( length(line) == 0) {
        break
      }
      m <- regmatches(line,regexec("^# (\\d+)\\. (.+)\\.$",line))[[1]]
      if (length(m) == 3) {
        index <- as.numeric(m[2])
        value <- m[3]
        if (is.function(fun.col)){
         cols <- fun.col(value,cols)
        }
        cols[index] <- value
        if (index == n.cols){
          break
        }
      }
    }
    close(con)
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=cols))
  }
  else{
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=seq(1:n.cols)))
  }
}

data <- read.pcibex("../course-materials/Files/results_FAKE_FOR_CHC9012.csv")
```

<br>

```diff
load("CHC9012_DataImportedFromPCIbex.RData")

view(data)
```
<br>

```{r}
#| echo: false
rmarkdown::paged_table(data, options = list(rows.print = 50))
```

---

## 👓 What are all these columns?

::::: panel-tabset
## Default columns from PCIbex

-  `Results.reception.time`: Time in the PCIbex server when the results were received. This can be used as the Participant ID, assuming that there are no two participants who submitted the results at the exact same time.
-  `MD5.hash.of.participant.s.IP.address`: Scripted IP address of the machine used by the participant.
-  `Controller.name`: Which controller was used to code the experiment.
-  `Order.number.of.item`: The order of presentation of the whole questionnaire, starting from the very first page.
-  `Inner.element.number`: Whether specific numbers were assigned to the element.
-  `Label`: The name given to the webpage in PCIbex (for example, `Welcome page`, `Instructions`, `Exeriment`, etc.).
-  `Latin.Square.Group`
-  `PennElementType`: The PCIbex element used for the action (`TextInput`, `Scale`, `Dropdown`, etc.).
-  `Parameter`: Breakdown of the elements used on a same page.
-  `Value`: What the participant wrote or selectedm or when an element *starts* or *ends*.
-  `EventTime`: Time point at which an action occured (the time corresponds to the PCIbex server)
-  `Comments`

## Columns that we manually logged

-  `ItemNo`: The ID of the item (here, sentence)
-  `Condition`: The experimental condition
-  `Group`
-  `UserAge`
-  `UserDegree`
:::::

---

## ✍️ Step 2: Preparatory activity {vertical-align="center"}

::: {.callout-tip icon="true"}
## Instructions

**What needs to be done to prepare the dataset?** 

Some tips:

-  Checking that it is complete and that nothing is missing,
-  Deleting the columns/rows that are not necessary, and 
-  Transforming (but not *altering*) the data
:::

---

## 💡 My own observations

::::: panel-tabset
## Missing data
-  [ ] Calculate the reaction time (so far, only the event time)
-  [ ] The actual sentences are missing, need to add them based on `ItemNo`
-  [ ] No information about the consent box, but this cannot be retrieved


## Unnecessary data

-  [ ] Columns containing the same or no value (`Controller.name`, `Inner.element.number`, `Latin.Square.Group`, `Comments`, for example)
-  [ ] Keep only the lines where `Label` is "Experiment"

## Data to transform

-  [ ] Columns to rename, to be more concise and explicit (`Results.reception.time`, `MD5.hash.of.participant.s.IP.address`)
-  [ ] Change the nature of the variable (`UserAge` to numerical value, `ItemNo` to a group/factor, etc.)

:::::


# 💻 Part 2: Arrange the dataset with RStudio, according to the plan

---

## 2.1 Remove unnecessary columns

-  We need to mention which column from which dataset. In R language, "the column called *Controller.name* from the dataset called *data*" is written using the `$` sign (`data$Controller.name`)
-  We assign `NULL` to this column to delete it.

```{r}
data$Controller.name <- NULL
data$Inner.element.number <- NULL
data$Latin.Square.Group <- NULL
data$Comments <- NULL
```

```r
data$Controller.name <- NULL
data$Inner.element.number <- NULL
data$Latin.Square.Group <- NULL
data$Comments <- NULL
```


## 2.2 Rename columns

::: {.r-stack}

::: {.fragment .fade-out fragment-index=1}
You can have a glimpse of the names of the columns using the `names(...)` function.

```r
names(data)
```
```{r}
names(data)
```
:::

::: {.fragment .fade-in-then-out fragment-index=1}
Each name is assigned to a numerical index. For instance, "[1]" corresponds to "Results.reception.time".

```r
names(data)[1]
```

```{r}
names(data)[1]
```
:::
::: {.fragment .fade-in fragment-index=2}
Once we have the index of columns, we just need to assign new names to them.

```r
names(data)[1] <- "ID_1_TimeReceived"
names(data)[2] <- "ID_2_IP"

names(data)
```
```{r}
names(data)[1] <- "ID_1_TimeReceived"
names(data)[2] <- "ID_2_IP"

names(data)
```

:::
:::

## 2.3 Select the rows we need

::: {style="display: grid; grid-template-columns: 1fr; grid-template-rows: 1fr; margin-top: 50px;"}
::: {.fragment .fade-out style="grid-area: 1 / 1;" fragment-index=1}

-  We use the `filter()` function from the `dplyr` package.
-  `%>%`: to say that "This is the `Label` column from the dataset called `data`".
-  `==` sign: we *want* the `Experiment` rows.
-  We store that into a new dataset called `data2`. 

```r
library(dplyr)

data2 <- data %>%
  filter(Label == "Experiment")
```
```{r}
#| message: false
#| warning: false

library(dplyr)

data2 <- data %>%
  filter(Label == "Experiment")
```
:::

::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=1}

**Check:** We expect to only find `Experiment` in the `Label` column.

```{r}
#| echo: false

rmarkdown::paged_table(data2, options = list(rows.print = 10))
```

:::
::: {.fragment .fade-in style="grid-area: 1 / 1;" fragment-index=2}

-  There are unnecessary `_Header_` rows in the `Parameter` column.
-  We can use the same logic to *remove* them, but we use `!=`, meaning "we *don't want* these rows".

```r
data2 <- data2 %>%
  filter(Parameter != "_Header_")
```
```{r}
data2 <- data2 %>%
  filter(Parameter != "_Header_")
```

```{r}
#| echo: false

rmarkdown::paged_table(data2, options = list(rows.print = 10))
```
:::
:::

## 2.4 Calculate the reaction time

::: {.callout-tip icon="false"}
## 📖 Defining reaction time

In this experiment, `Reaction time` = `Time when the judgment is made` - `Time when the sentence appeared`

Calculating the reaction time from PCIbex data requires several steps.
:::

Let's have a look at our data:

```{r}
#| echo: false
data_show <- data2 %>%
  select(Parameter, Value, EventTime, ItemNo)
rmarkdown::paged_table(data_show, options = list(rows.print = 6))
```

Here, `Reaction time` = "EventTime *when* `Parameter` = `Choice`" - "EventTime *when* `Value` = `Start`". So we need to select these rows first. We stored in a temporary dataset that we call `RT_Prep1`.

```{r}
RT_Prep1 <- data2 %>%
  filter(Parameter == "Choice" | Value == "Start")
```

Under the `Value` column, we have the time for `Start`, but not for `Choice`, which is found in another column. We want to have everything under the same column. So we tell R that when in the `Value` column, the word is not `Start`, it writes `Choice` instead. Then, we delete columns when the two rows are different -- we want to keep the differences minimal.

```{r}
RT_Prep1$Value <- ifelse(RT_Prep1$Value == "Start",
                         "Start",
                         "Choice")

RT_Prep1$PennElementType <- NULL
RT_Prep1$PennElementName <- NULL
RT_Prep1$Parameter <- NULL
```

```{r}
#| echo: false

RT_Show <- RT_Prep1 %>%
  select(Value, EventTime, ItemNo)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

To do the subtraction, we want to have the `EventTime` for `Start` and `Choice` in different columns. We use the `spread()` function from the `tidyr` library.

```{r}
library(tidyr)
RT_Prep2 <- spread(data = RT_Prep1, Value, EventTime)

```

```{r}
#| echo: false

RT_Show <- RT_Prep2 %>%
  select(ItemNo, Start, Choice)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

Now we just have to tell R that we are dealing with numbers, and do the subtraction.

```{r}
RT_Prep2$Choice <- as.numeric(RT_Prep2$Choice)
RT_Prep2$Start <- as.numeric(RT_Prep2$Start)
RT_Prep2$RT <- RT_Prep2$Choice - RT_Prep2$Start
```

```{r}
#| echo: false

RT_Show <- RT_Prep2 %>%
  select(ItemNo, Start, Choice, RT)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 6))
```

::: {.callout-tip icon="false"}
## 💭 Last step

We have two datasets:

-  `data2`, with the judgment data, but ***not*** the reaction time,
-  `RT_Prep2`, the temporary dataset with the reaction time, but ***not*** the judgment data.

We just need to join them based on their common minimal value: 1 different trial per participant.
:::

First, we select the row where the judgment is given.

```{r}
data3 <- data2 %>%
  filter(Parameter %in% "Choice")
```

Finally, we use the `left_join()` function to join these two datasets based on (a) the participant ID, and (b) the Item index.

```{r}
data3 <- data3 %>%
  left_join(RT_Prep2 %>% select(ID_1_TimeReceived, ID_2_IP, ItemNo, RT), 
            by = c("ID_1_TimeReceived", "ID_2_IP", "ItemNo"))
```
```{r}
#| echo: false

RT_Show <- data3 %>%
  select(ID_1_TimeReceived, ItemNo, Value, RT, Condition)
rmarkdown::paged_table(RT_Show, options = list(rows.print = 25))
```


## 2.5 Change the nature of the variables

::: {style="display: grid; grid-template-columns: 1fr; grid-template-rows: 1fr; margin-top: 50px;"}
::: {.fragment .fade-out style="grid-area: 1 / 1;" fragment-index=1}
Let's look again at our data:
```r
str(data3)
```
```{r}
str(data3)
```
:::

::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=1}
Some variables correspond to ***groups*** (or categories), so we used the `as.factor()` function to group them.

```r
data3$ID_1_TimeReceived <- as.factor(data3$ID_1_TimeReceived)
data3$ID_2_IP <- as.factor(data3$ID_2_IP)
data3$ItemNo <- as.factor(data3$ItemNo)
data3$Condition <- as.factor(data3$Condition)
data3$Group <- as.factor(data3$Group)
data3$UserDegree <- as.factor(data3$UserDegree)
```
```{r}
data3$ID_1_TimeReceived <- as.factor(data3$ID_1_TimeReceived)
data3$ID_2_IP <- as.factor(data3$ID_2_IP)
data3$ItemNo <- as.factor(data3$ItemNo)
data3$Condition <- as.factor(data3$Condition)
data3$Group <- as.factor(data3$Group)
data3$UserDegree <- as.factor(data3$UserDegree)
```
:::

::: {.fragment .fade-in style="grid-area: 1 / 1;" fragment-index=2}
For numerical values, we use the `as.numeric()` function.

```r
data3$Order.number.of.item <- as.numeric(data3$Order.number.of.item)
data3$Value <- as.numeric(data3$Value)
data3$UserAge <- as.numeric(data3$UserAge)
data3$RT <- as.numeric(data3$RT)
```
```{r}
data3$Order.number.of.item <- as.numeric(data3$Order.number.of.item)
data3$Value <- as.numeric(data3$Value)
data3$UserAge <- as.numeric(data3$UserAge)
data3$RT <- as.numeric(data3$RT)
```

Now this is our data:
```r
str(data3)
```
```{r}
str(data3)
```
:::
:::

## 2.6 Another round of removing unnecessary columns

::: {style="display: grid; grid-template-columns: 1fr; grid-template-rows: 1fr; margin-top: 50px;"}
::: {.fragment .fade-out style="grid-area: 1 / 1;" fragment-index=1}
We made use of some columns when calculating the reaction time, but we do not need them anymore, and they are redundant. So here is another round of column removal!

```r
str(data3)
```
```{r}
str(data3)
```
:::
::: {.fragment .fade-in style="grid-area: 1 / 1;" fragment-index=1}

```r
data3$Label <- NULL
data3$PennElementType <- NULL
data3$PennElementName <- NULL
data3$Parameter <- NULL
data3$EventTime <- NULL

str(data3)
```
```{r}
data3$Label <- NULL
data3$PennElementType <- NULL
data3$PennElementName <- NULL
data3$Parameter <- NULL
data3$EventTime <- NULL

str(data3)
```
:::
:::

## 2.7 Add information about the items

Finally, we forgot to `.log()` the sentences in the PCIbex script:

-  We import the table with the material information (used on the PCIbex platform),
-  We use the `left_join()` function to join these two datasets based on the index of the items (here, `ItemNo`)

```r
material <- read.csv("~/Documents/CHC9012_Spring2026/PreprocessingAnalysesR/material.csv")
material$ItemNo <- as.factor(material$ItemNo)

data3 <- data3 %>%
  left_join(material %>% select(ItemNo, Sentence), 
            by = "ItemNo")

data3$Sentence <- as.factor(data3$Sentence)
str(data3)
```
```{r}
material <- read.csv("~/Documents/CHC9012_Spring2026/PreprocessingAnalysesR/material.csv")
material$ItemNo <- as.factor(material$ItemNo)

data3 <- data3 %>%
  left_join(material %>% select(ItemNo, Sentence), 
            by = "ItemNo")

data3$Sentence <- as.factor(data3$Sentence)
str(data3)
```


# 💾 Part 3: Check and save the data

---

## 💭 Checking process

::: {.callout-tip icon="true"}

We want to make sure that we did not corrupt our data when arranging them for further analyses. Otherwise, we can run further analyses and have results, but they will have no value!

We will mostly use the `ezDesign()` function from the `ez` package.
:::

---

## 3.1 Check the data

::: {style="display: grid; grid-template-columns: 1fr; grid-template-rows: 1fr; margin-top: 50px;"}
::: {.fragment .fade-out style="grid-area: 1 / 1;" fragment-index=1}

We load the library:

```r
library(ez)
```
```{r}
#| message: false
#| warning: false
library(ez)
```
:::
::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=1}

```r
# Same number of item per condition for each participant? (should be '15')
ezDesign(data3, ID_1_TimeReceived, Condition)
```
```{r}
# Same number of item per condition for each participant? (should be '15')
ezDesign(data3, ID_1_TimeReceived, Condition)
```
:::
::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=2}
```r
# Which participants share the same ID? (hint that this could be the same participant filling the questionnaire twice)
ezDesign(data3, ID_1_TimeReceived, ID_2_IP)
```
```{r}
# Which participants share the same IP?
# (hint that this could be the same participant filling the questionnaire twice)
ezDesign(data3, ID_1_TimeReceived, ID_2_IP)
```
:::
::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=3}
```r
# Number of times an item appeared by participant (should be '1')
ezDesign(data3, ID_1_TimeReceived, ItemNo)
```
```{r}
# Number of times an item appeared by participant (should be '1')
ezDesign(data3, ID_1_TimeReceived, ItemNo)
```
:::
::: {.fragment .fade-in-then-out style="grid-area: 1 / 1;" fragment-index=4}
```r
# Were the items well dispatched in the two lists?
ezDesign(data3, ItemNo, Group)
```
```{r}
# Were the items well dispatched in the two lists?
ezDesign(data3, ItemNo, Group)
```
:::
::: {.fragment .fade-in style="grid-area: 1 / 1;" fragment-index=5}
```r
# Were the participants well dispatched in the two lists? (should be '30')
ezDesign(data3, ID_1_TimeReceived, Group)
```
```{r}
# Were the participants well dispatched in the two lists? (should be '30')
ezDesign(data3, ID_1_TimeReceived, Group)
```
:::
:::

## 3.2 Save the prepared dataset

```r
save(data3,
     file = "CHC9012_PCIbexDataPrepared.RData")
```
```{r}
save(data3,
     file = "CHC9012_PCIbexDataPrepared.RData")
```

Check that the file appears in your working directory!


# 📝 Assignment for next week

::: {.callout-important icon="true"}
Please annotate the Markdown file from this week, knit to a PDF file, and upload on Moodle!
:::
